\documentclass{tufte-handout}
\usepackage{amsmath,amsthm}

\input{vc.tex}

\usepackage{pgfplots}
\pgfplotsset{width=\textwidth}

\newtheorem{claim}{Claim}[section]
\title{\sf Exact Algorithm for Independent Set}
\date{\GITAuthorDate, rev. \GITAbrHash}
\author{}

\newcommand{\mis}{\textsc{MIS}}
\newcommand{\nodeu}{\ensuremath{\tiny u}}
\newcommand{\nodev}{\ensuremath{\tiny v}}

\tikzset{
  auto,node distance=5mm,thick,
  remaining node/.style={thick,circle,draw,minimum size=3mm,fill=black!10},
  remaining edge/.style={thick},
  mis node/.style={dotted,circle,draw,minimum size=3mm,fill=blue!20},
  removed node/.style={dotted,circle,draw,minimum size=3mm},
  removed edge/.style={thin,dotted}
}

\begin{document}
\maketitle

\section{Independent Set}
Consider an undirected unweighted graph $G=(V,E)$ and set $n=|V|$.
The Maximum Independent Set problem (MIS) is to find a subset $S$ of
the vertices of largest size such that no pair of different vertices
in $S$ are connected by an edge in $G$.
The size of the largest independent set in $G$ is denoted by
$\alpha(G)$.

MIS is NP-hard, but we will nevertheless implement algorithms that
compute $\alpha(G)$ for small instance graphs $G$ and argue both
empirically and theoretically about their running time.

\paragraph{Notation.}
For a subset $X\subseteq V$, we will by $G[X]$ denote the graph
\emph{induced} by $X$, the graph that remains if we keep only the
vertices in $X$ and the edges among them.
For a vertex $v$, we denote by $N[v]$ its \emph{closed neighborhood},
i.e. $v$ and all its neighbors.
We will slightly abuse set difference and union notation by using $-$
and $+$ instead.
$X+Y-Z$ for three sets $X,Y,Z$ are those elements that are in either
$X$ or $Y$ but not in $Z$.


\subsection{Algorithm $R_0$}

Consider the following simple recursive algorithm working on an input
graph $G=(V,E)$.
(Call it Algorithm~$R_0$):
\begin{enumerate}
\item If the input graph is empty, return $0$.
\item If the input graph $G$ has a vertex $v$ without any neighbors,
  return $1+R_0(G[V-v]).$ See fig.~1.

  \begin{marginfigure}
\begin{tikzpicture}
\node (a) at (0,0){
      \begin{tikzpicture}
      \node[remaining node] (1) {};
      \node[remaining node] (2) [above left of=1] {};
      \node[remaining node] (3) [above right of=1] {};
      \node[remaining node] (4) [below left of=1] {};
      \node[remaining node] (5) [below of=1] {};
      \node[remaining node] (6) [below right of=1] {};
      \node[remaining node] (7) [above right of=6] {};
      \node[remaining node] (8) [above of=7] {};
      \node[remaining node] (9) [above right of=2] {\nodev};

      \path[remaining edge]
      (1) edge node {} (2)
          edge node {} (3)
          edge node {} (4)
          edge node {} (5)
          edge node {} (6)
      (4) edge node {} (5)
      (6) edge node {} (7)
      (7) edge node {} (8);
    \end{tikzpicture}};
\node (b) at (2,0) {
    \begin{tikzpicture}
      \node[remaining node] (1) {};
      \node[remaining node] (2) [above left of=1] {};
      \node[remaining node] (3) [above right of=1] {};
      \node[remaining node] (4) [below left of=1] {};
      \node[remaining node] (5) [below of=1] {};
      \node[remaining node] (6) [below right of=1] {};
      \node[remaining node] (7) [above right of=6] {};
      \node[remaining node] (8) [above of=7] {};
      \node[mis node] (9) [above right of=2] {\nodev};

      \path[remaining edge]
      (1) edge node {} (2)
          edge node {} (3)
          edge node {} (4)
          edge node {} (5)
          edge node {} (6)
      (4) edge node {} (5)
      (6) edge node {} (7)
      (7) edge node {} (8);
    \end{tikzpicture}
  };
    \draw[->, very thick] (a) -- (b);
 \end{tikzpicture}
    \caption{A graph with an isolated node $v$.
      We can safely say $v$ is in the MIS, and continue without it.}
  \end{marginfigure}

\item Otherwise find a vertex $u$ of maximum degree (if there are
  several any will do) and return 
\[\max(1+R_0(G[V-N[u]]),
  R_0(G[V-u]))\,,\]keeping in mind that that $N[u]$ includes $u$. See fig.~2.

  \begin{marginfigure}
    \begin{tikzpicture}
\node (base) at (0,0) {
    \begin{tikzpicture}
      \node[remaining node] (1) {\nodeu};
      \node[remaining node] (2) [above left of=1] {};
      \node[remaining node] (3) [above right of=1] {};
      \node[remaining node] (4) [below left of=1] {};
      \node[remaining node] (5) [below of=1] {};
      \node[remaining node] (6) [below right of=1] {};
      \node[remaining node] (7) [above right of=6] {};
      \node[remaining node] (8) [above of=7] {};
      \node[remaining node] (9) [above right of=2] {};

      \path[remaining edge]
      (1) edge node {} (2)
          edge node {} (3)
          edge node {} (4)
          edge node {} (5)
          edge node {} (6)
      (4) edge node {} (5)
      (6) edge node {} (7)
      (3) edge node {} (9)
          edge node {} (8)
      (2) edge node {} (9);
      \path[remaining edge]
      (7) edge node {} (8);
    \end{tikzpicture}};
    \node (left) at (-1.5,-2){
    \begin{tikzpicture}
      \node[mis node] (1) {\nodeu};
      \node[removed node] (2) [above left of=1] {};
      \node[removed node] (3) [above right of=1] {};
      \node[removed node] (4) [below left of=1] {};
      \node[removed node] (5) [below of=1] {};
      \node[removed node] (6) [below right of=1] {};
      \node[remaining node] (7) [above right of=6] {};
      \node[remaining node] (8) [above of=7] {};
      \node[remaining node] (9) [above right of=2] {};

      \path[removed edge]
      (1) edge node {} (2)
          edge node {} (3)
          edge node {} (4)
          edge node {} (5)
          edge node {} (6)
      (4) edge node {} (5)
      (3) edge node {} (9)
          edge node {} (8)
      (6) edge node {} (7)
      (2) edge node {} (9);
      \path[remaining edge]
      (7) edge node {} (8);       
    \end{tikzpicture}};
    \node (right) at (1.5,-2) {
    \begin{tikzpicture}
      \node[removed node] (1) {\nodeu};
      \node[remaining node] (2) [above left of=1] {};
      \node[remaining node] (3) [above right of=1] {};
      \node[remaining node] (4) [below left of=1] {};
      \node[remaining node] (5) [below of=1] {};
      \node[remaining node] (6) [below right of=1] {};
      \node[remaining node] (7) [above right of=6] {};
      \node[remaining node] (8) [above of=7] {};
      \node[remaining node] (9) [above right of=2] {};

      \path[removed edge]
      (1) edge node {} (2)
          edge node {} (3)
          edge node {} (4)
          edge node {} (5)
          edge node {} (6);
      \path[remaining edge]
      (4) edge node {} (5)
      (6) edge node {} (7)
      (7) edge node {} (8)
      (3) edge node {} (9)
          edge node {} (8)
      (2) edge node {} (9);
    \end{tikzpicture}};
  \draw [->, very thick] (base) -- (left);
  \draw [->, very thick] (base) -- (right);
  \end{tikzpicture}

    \caption{Vertex $u$ has highest degree.
      Left branch: Assume $u$ is in the MIS -- remove $u$ and its
      neighbors, since the neighbors can no longer be in the MIS.
      Right branch: Assume $u$ not in the MIS -- just remove $u$.}
  \end{marginfigure}
\end{enumerate}
To see why algorithm $R_0$ correctly computes $\alpha(G)$, note that
if there is an isolated vertex $v$, a maximum size independent set
will always contain it (Row 2).
On the other hand, if there are no isolated vertices, we can pick a
vertex $u$ of maximum degree and branch on the two cases that $u$ is
in the maximum independent set (in which case we discard all vertices
adjacent to $u$ from $G$), or that $u$ is not in the maximum
independent set (Row 3).

\subsection{Inputs}

The data directory contains eleven random input instances of
increasing size. 
All input files are text files on the following format describing the
undirected graph $G=(V,E)$: First comes a positive integer $n$ giving
the number of vertices in the graph.
Next follow $n$ rows each containing $n$ $0/1$-entries describing the
so-called \emph{adjacency matrix} of the graph.
The $j$th integer at the $i$th row is a $1$ if and only if $ij\in E$.

\begin{marginfigure}
\begin{tabular}{lrr}
file & $|V|$ & $\alpha(G)$\\
 \texttt{g30.in} & 30 & 14\\
 \texttt{g40.in} & 40 & 16\\
 \texttt{g50.in} & 50 & 19\\
 \texttt{g60.in} & 60 & 20\\
 \texttt{g70.in} & 70 & 22\\
 \texttt{g80.in} & 80 & 24\\
 \texttt{g90.in} & 89 & 26\\
 \texttt{g100.in} & 100 & 27\\
 \texttt{g110.in} & 110 & 29\\
 \texttt{g120.in} & 120 & 30\\
 \texttt{g130.in} & 130 &  ? 
\end{tabular}
\end{marginfigure}

\subsection{Algorithm $R_1$}
Add the following line to algorithm $R_0$ after its first line
checking for emptiness, and call the resulting algorithm $R_1$
(replacing the recursive calls to $R_0$ by $R_1$):

If $G$ has a vertex $v$ of degree $1$ return $1+R_1(G[V-N[v]])$.

Before you write any code -- draw (by hand) pictures such as those
above to illustrate what happens with your graph when you run
algorithm $R_1$ on it. Use the pictures when you implement the
algorithm, and when you argue the correctness of it, i.e., when you
motivate why it always computes $\alpha(G)$. You are supposed to
bring your pictures and show them to your lab assistant.

%Algorithm $R_1$ is just a small modification to Algorithm $R_0$. 

%The difference is just to also inspect nodes $u$ with only one neigbour. 
%if there exists such a node $u$, just return $1 + R_1(G[V-N[u]])$.

\subsection{Algorithm $R_2$}
Add the following line to algorithm $R_1$ after its first line
checking for emptiness, and call the resulting algorithm $R_2$
(replacing the recursive calls to $R_1$ by $R_2$):

If $G$ has a vertex $v$ with exactly two neighbors $u,w$, then if
$uw\in E$ return $1+R_2(G[V-N[v]])$, else add a new vertex $z$ to
the graph with edges to every neighbor of $u$ and $w$, except $v$,
and return $1+R_2(G[V-N[v]+z])$.

If $G$ doesn't have a vertex $v$ with exactly two neighbors, then we
proceed just as in algorithm $R_1$.

Draw pictures such as those above to illustrate algorithm $R_2$, and
use the pictures both when you implement it, and when you argue the
correctness of it, i.e., motivate why it always computes
$\alpha(G)$.

\subsection{Deliverables}

\begin{enumerate}
\item Implement algorithm $R_0$ and run it on the instances provided in
  data/g30.txt, data/g40.txt, ..., data/g60.txt.
  
  Count the number of recursive calls of $R_0$ for each instance and plot the
  logarithm of that number vs the instance vertex size. 
  
  What is the time complexity dependence on $n$?
  
  Use whatever programming language and libraries you want, but make
  sure that your code is not unnecessarily long.

  First implement the algorithm by making a new copy of the graph at
  each recursive call, just to make it as easy as possible to verify
  that the algorithm works as intended. Once you get this to work, you
  can pay attention to the problem of efficiently representing a graph
  through the recursion so that it is easy to remove (and add, see
  below) vertices to it by temporarily modifying the graph at hand.

\item After you have implemented algorithm $R_1$, run it on the instances
 data/g30.in, data/g40.in, $\ldots$, data/g100.in.

  Count the number of recursive calls of $R_1$ for each instance and plot the
  logarithm of that number vs the instance vertex size.
  
  What is the time complexity dependence on $n$?

\item Run algorithm $R_2$ on the instances data/g30.in, data/g40.in,
  $\ldots$, data/g120.in.

  Count the number of recursive calls of $R_2$ for each instance and plot the
  logarithm of that number vs the instance vertex size. 
  
  What is the time complexity dependence on $n$?

\item Fill out the report on the next page; you can just use the
  \LaTeX\ code if you want.
\end{enumerate}

\section{Optional}
Solve independent set using bitmasks. This is a good exercise to prepare for the last lab Tree Width.

\subsection{Algoritm $MM$}
We will use a trick called \emph{meet in the middle}, which can be used on several NP-complete problems.

\bigskip

Meet in the middle is similar to divide and conquer. We split the problem into two halves, solve them seperately and then merge the results.
However, we can't continue to divide recursively, because the subproblems that are solved in each half are not symmetric, 
and also differ slightly to the original problem.

\bigskip


For independent set, we divide the input graph $G$ in two halves $A$ and $B$.
We define a function $f: B' \rightarrow int$, where $B'$ is a subset of $B$. 
$f(B')$ is the size of the largest subset of $B'$ that is independent.

\bigskip

The subsets $A'$ of $A$ that are independent are saved together with their size.

Then we iterate through all the saved subsets $A'$ and compute 
the maximal independent set of $G$ where exactly the nodes $A'$ from $A$ are used, by querying $B$ in the following way.
If we want to use exactly the nodes $A'$ from $A$ we can't use any of the neighbors of $A'$ in $B$. 
We can therefor ask $f$ of the maximal independent set of the nodes $B' = B - N[A']$, see figure \ref{fig:MM}. 
The maximal size of an independent set $A' \cup B'$ is then $f(B') + |A'|$.
The maximal independent set of $G$ is $max_{A'}(f(B') + A')$ over all $A'$,
since this search is exhaustive over $A$.

  \begin{marginfigure}
  \label{fig:MM}
\begin{tikzpicture}
\node (a) at (0,0){
      \begin{tikzpicture}
          \node[mis node] (1) at (0, 0) {\ensuremath{\tiny A0}};
          \node[remaining node] (2) at (0.7, -0.5) {\ensuremath{\tiny A1}};
          \node[remaining node] (3) at (0, 1) {\ensuremath{\tiny A2}};
          \node[mis node] (4) at (0.7, 1.5) {\ensuremath{\tiny A3}};
          \node[remaining node] (5) at (1.8, -0.5) {\ensuremath{\tiny B0}};
          \node[remaining node] (6) at (2.3, 0) {\ensuremath{\tiny B1}};
          \node[remaining node] (7) at (2.3, 1) {\ensuremath{\tiny B2}};
          \node[remaining node] (8) at (1.8, 1.5) {\ensuremath{\tiny B3}};

      \path[remaining edge]
      (1) edge node {} (2)
          edge node {} (3)
          edge node {} (6)
          edge node {} (7)
      (2) edge node {} (5)
      (4) edge node {} (8)
      (6) edge node {} (7)
      (7) edge node {} (8);
    \end{tikzpicture}};
\node (b) at (0,-4) {
     \begin{tikzpicture}
          \node[mis node] (1) at (0, 0) {\ensuremath{\tiny A0}};
          \node[remaining node] (2) at (0.7, -0.5) {\ensuremath{\tiny A1}};
          \node[remaining node] (3) at (0, 1) {\ensuremath{\tiny A2}};
          \node[mis node] (4) at (0.7, 1.5) {\ensuremath{\tiny A3}};
          \node[remaining node] (5) at (1.8, -0.5) {\ensuremath{\tiny B0}};
          \node[removed node] (6) at (2.3, 0) {\ensuremath{\tiny B1}};
          \node[removed node] (7) at (2.3, 1) {\ensuremath{\tiny B2}};
          \node[removed node] (8) at (1.8, 1.5) {\ensuremath{\tiny B3}};

      \path[remaining edge]
      (1) edge node {} (2)
          edge node {} (3)
          edge node {} (6)
          edge node {} (7)
      (2) edge node {} (5)
      (4) edge node {} (8)
      (6) edge node {} (7)
      (7) edge node {} (8);

    \end{tikzpicture}};
    \draw[->, very thick] (a) -- (b);
 \end{tikzpicture}
    \caption{A graph where $A'$ is chosen as nodes $\{ A0, A3 \}$. Now $B' = B - N[A'] = \{ B0\} $, since $\{ B1, B2, B3 \} $ all are connected to $A'$.}
  \end{marginfigure}
\bigskip

This can be efficiently implemented using bitstrings and memoization over all the $2^{|B|}$ inputs to $f$. 
$f$ can either be implemented recursively, or by just filling in a table.

\bigskip

Note that $f$ can be defined recursively in the following way: 
$f(B') = |B'| \texttt{ if }B' \texttt{ is independent, else } max_{b \in B'}f(B' - b)$.

\bigskip

Given that $|B| < 32$, we can encode a subset of $B$ as a 32-bit integer.
We label all the nodes in $B$ as $0, 1, 2, \ldots, |B| - 1$. Then to compute the bitstring (integer) of a subset,
we sum the powers of two of the labels.

For example, if we want to find the bitstring of the subset containing nodes with labels $\{ 0, 3, 4, 6 \} $ we compute the bitstring as $2^0 + 2^3 + 2^4 + 2^6 = 89$.
In most programming languages we can use $1\ll x$ to compute $2^x$.

\subsection{Discussion topics for algorithm MM}
\begin{itemize}
    \item What is the running time?
    \item Can you find a graph where MM is faster than R1?
    \item Implementation complexity compared to R2.
\end{itemize}

\newpage


\newpage
\section{Independent Set Lab Report}


by Alice Cooper and Bob Marley\sidenote{Complete the report by filling
  in your names and the parts marked $[\ldots]$.
  Remove the sidenotes in your final hand-in.}

\subsection{Correctness}
Algorithm R1 correctly computes $\alpha(G)$ because $[\ldots]$.\sidenote{Fill in the $[\dots]$ with an argument why R1 indeed computes the size of the maximum independent set. Use as little mathematical notation as necessary. A ``proof'' is whatever convinces me.}

\noindent
Algorithm R2 correctly computes $\alpha(G)$ because $[\ldots]$.

\subsection{Empirical Running time}

\paragraph{Experiments.}
\sidenote{Plot the logarithm of the number of recursive calls for the
  three algorithms $R_0$, $R_1$, and $R_2$, in a combined image.
  
  Use whatever software you like to produce the image; the placeholder
  image on the left is constructed in the \LaTeX\ source, and should
  give you a feel of what the result could look like (with completely
  different values on the y-axis.)}

\medskip

\noindent
\begin{tikzpicture}
\begin{axis}[
  height= 5cm,
  xlabel=$n$,
  ylabel={log recursive calls},
  xmin = 20,  xmax = 150,
  xtick =       {20,40,60,80,100,120},
  xticklabels = { $20$, $40$, $60$, $80$, $100$, $120$},
  x tick label as interval = false,
  scaled ticks = false
]
    \addplot[color=black, mark=x] 
	coordinates {
	(30, 100)
	(40, 200)
	(50, 300)
	(60, 400)
    };
    \addlegendentry{R0}

    \addplot[color=black, mark=o] 
	coordinates {
	(30, 50)
	(40, 100)
	(50, 150)
	(60, 200)
	(70, 250)
	(80, 300)
	(90, 350)
	(100, 400)
    };
    \addlegendentry{R1}

    \addplot[color=black, mark=*] 
	coordinates {
	(30, 30)
	(40, 60)
	(50, 90)
	(60, 120)
	(70, 150)
	(80, 180)
	(90, 210)
	(100, 240)
	(110, 270)
	(120, 300)
    };
    \addlegendentry{R2}

\end{axis}
\end{tikzpicture}

The running times of algorithm~$R_0$, $R_1$, and $R_2$ appear to be
$[\ldots],[\ldots],$ and $[\ldots]$, respectively.\sidenote{Replace
  the $[\ldots]$ by a function of $n$ on the form $O(c^n)$.
  Use your measurements on the run time to estimate $c$ in the three
  cases.}.

\subsection{Theoretical Upper Bound}

Denote be $T_i(n)$ the worst runtime of algorithm Ri on \emph{any} graph on $n$ vertices.
Note that $T_i(n)$ is a non-decreasing function of $n$.
For $R_0$ we can conclude that
\begin{align*}
T_0(n) &\leq\max(T_0(n-1), T_0(n-1)+T_0(n-1-d_{\mbox{max}})) \\ &\leq T_0(n-1)+T_0(n-2)
\end{align*}
with $d_{\mbox{max}}$ the degree of the vertex we branch on. The hard part is the one when there are no isolated vertices, in which case the vertex $u$ we are branching on has at least one neighbor. 

For R1 we have that\sidenote{Derive similar recursive bounds on $T_1(n)$ and $T_2(n)$.}
 \[
 T_1(n)=[\ldots]
 \]

For R2 we have that
 \[
 T_2(n)=[\ldots]
 \]
\paragraph{Worst Case Upper Bound}
The running times of algorithm~$R_0$, $R_1$, and $R_2$ are in
$[\ldots],[\ldots],$ and $[\ldots]$, respectively.\sidenote{Replace
  the $[\ldots]$ by a function of $n$ on the form $O(c^n)$.
  Use the recursive bounds you've derived above.
  \emph{Hint: A recurrence of the form $T(n)\leq \sum_{i=1}^k
    a_iT(n-i)$ is called a linear homogeneous recurrence relation with
    constant coefficients.
    To solve it, you can set $T(n)\leq c^n$ where $c$ is the largest
    real root to the characteristic polynomial $x^k-\sum_{i=1}^k
    a_ix^{k-i}$}.}. \newpage


\newpage
\section{ Perspective}

This lab establishes medium skills in recursive algorithms implementation,
and simple means to analyze their running time.

\bigskip

The algorithms investigated here can hardly be called ``advanced'', but the main message 
is that no one knows how to do significantly better with any other means.
Indeed, the strongest theoretical worst case run time bound for maximum independent set to date
is obtained by a computerized calculation on a huge set of branching reduction rules, just like the ones we've looked at here [Robson. Algorithms for maximum independent sets. Journal of Algorithms, 7(3):425--440, 1986]. For a much cleaner analysis with just a few rules, obtaining a comparative bound, see [Fomin, Grandoni, and Kratsch. A measure \& conquer approach for the analysis of exact algorithms, JACM 56 (5), article No. 25, 2009].


\end{document}
